\sysfull

### 2017-08-08 paper/paging.tex
Small corrections (typos/capitals) to paging

\subsection{Paged non-volatile virtual memory}

\sys uses a page-based virtual memory scheme that allows tasks to primarily use
SRAM instead of frequently accessing FRAM.  The paging scheme partitions the
non-volatile memory address space into pages of {\em protected data}. As they
are accessed by a task, pages of protected data are swapped into a {\em page
buffer} that is stored in volatile memory and holds a single page. When a
modified page is swapped out, it is buffered in a dedicated commit buffer for
later commit.  \sys uses two-phase commit to write buffered, modifed pages back
to their original memory locations atomically, ensuring data consistency even
with power failures.
%
\sys's page-based memory model allows moving large blocks of data between
non-volatile and volatile memory, which \sys does efficiently using common,
hardware DMA block-copy operations that have higher bandwidth and avoid fixed,
per-variable data movement costs.  \sys uses efficient DMA block copies when
paging data in and out, as well as during two-phase commit.  \sys is unlike
prior work~\cite{chain,alpaca} that moves one variable at a time, prohibiting
efficient, DMA bulk transfers. 
%
Moreover, \sys's paging and buffering scheme is a key enabler of task
coalescing, which we describe in Section~\ref{sec:task_coalescing}. Paging
simplifies coalescing because deferring a commit of paged data at a task
boundary entails leaving paged data buffered only, not requiring
additional state management logic at task boundaries.

@@

\sys virtualizes the protected, non-volatile data that a task can access
through a modified memory interface.  The interface includes two operations:
\texttt{RVAR(var)}, which reads the value of the memory location {\tt var} and
\texttt{WVAR(var,val)}, which writes the value {\tt val} into the memory
location {\tt var}.  
%
{\tt RVAR} and {\tt WVAR} operations translate a variable's physical address in
non-volatile memory and translates it into a \emph{virtual address} in the page
buffer \texttt{pageBuf} in volatile memory.  In \sys, a virtual address is
composed of a \emph{page tag} and a \emph{page offset}.  The page tag
identifies in which page of non-volatile memory the memory location resides and
the page offset identifies a particular byte of that page.  
%
How an $a$-bit memory address decomposes into a tag and an offset depends on
the number (and size) of each page of protected memory.  To address {\tt p}
pages of size {\tt s}, a tag comprises $log(p)$ bits, the offset comprises
$log(s)$ bits and $log(p) + log(s) = a$.   For example, in a 16-bit address
space (i.e., 16kB of protected memory) with 512 pages of 128 bytes each, a
page's tag is the high-order nine bits of the address and a byte's page offset
is the remaining 7 bits.

@@

After address translation, \sys attempts to access the protected variable's
location in the volatile paging buffer.  \sys keeps track of the page tag for
the page currently resident in the paging buffer in {\tt CrntPagTag}, which is
stored in non-volatile memory. When a task accesses a variable, it compares the
variable's page tag to {\tt CrntPagTag}.  If an accessed variable's page is
equal to {\tt CrntPagTag} then the accessed variable's page is resident in the
page buffer.  If the accessed variable's page tag is not equal to {\tt
CrntPagTag} then the variable's page is not resident in the page buffer.
%
If the accessed variable's page is resident in the page buffer, an executing
memory operation indexes the page buffer using the variable's page offset and
manipulates that byte (i.e., either reading or writing).
%
If the accessed variable's page is not resident in the page buffer, the
operation incurs a {\em page fault}. At a page fault, \sys must swap out the
page that is resident int he page buffer and swap in the the variable's page.

Algorithm~\ref{algo:rwar} shows pseudo-code of \texttt{RVAR} and we omit a
listing for {\tt WVAR} because it is very similar except it writes the memory
location and returns nothing.  If the page tag of the variable \texttt{var} and
\texttt{CrntPagTag} are not equal (e.g. Lines 2--3), there is a page fault
that requires a new page to be swapped in to the page buffer. 
%Observe that the computational complexity of the whole steps pertaining to
%address translation and variable access is $\mathcal{O}(1)$---introducing
%minimal overhead to the system to boost up benefits of volatile-only memory
%interfacing.

@@

A page fault on a memory access requires \sys to swap out whatever page of data
is in the page buffer, preserving updates made to that page, and to copy the
accessed page of protected data into the page buffer. 
%
Algorithm~\ref{algo:pagefault} shows the how \sys handles a page fault.  If
both the page being swapped out and the page being swapped in are not {\em
dirty} --- which means that they do not contain bytes modified by the executing
task --- then handling a page fault simply requires copying the accessed page
into the page buffer.  If either the {\em resident} page being swapped out or
the {\em accessed} page being swapped in is dirty, \sys must take extra steps
when handling a page fault.

@@

\paragraph{Dirty accessed page.} If an accessed page is modified, \sys should
swap it into the page buffer from its location in {\tt pagesTemp}, rather than
swapping it in from its original location in memory. 
%
As Algorithm~\ref{algo:pagefault} shows in Line 3, \sys checks if the requested
page was previously modified by the executing task. To track whether a page is
modified, \sys stores a dirty bit with the page that is set when a task
modifies the page.  If an accessed page's dirty bit is set, the page must be
swapped in from \texttt{pagesTemp} rather than its original memory location
(Line 4), furnishing the task with the page's modified values. If the accessed
page is not dirty, \sys swaps it in from its original location in memory (Line
6). After swapping the page in, \sys sets \texttt{CrntPagTag} to the tag of the
accessed page (Line 7). \sys uses DMA block copies for all page swapping
operations, leveraging the efficiency of hardware support.

\subsubsection{Committing a page}

@@

\subsection{Paging Example}
\todo{Redo this figure to actually reflect paging -- right now it isn't clear if x,y,z are in a page or not, it isn't clear when the page gets copied to pagesTemp, it isn't clear what state we add to the system to help with commit, etc (i.e., dirty bits, commit bit).  Make
this figure's labels consistent with the text: pagesTemp, RVAR, WVAR, explicit
commit in task, Page Buffer}{whoever} 
Figure~\ref{fig:volatile-buffer} illustrates the execution of a simple task
using \sys's paging mechanism.  The task manipulates protected variables
{\tt x}, {\tt y}, and {\tt z}, which are in the same page.  On accessing {\tt
x}, its page is swapped into the page buffer.  On accessing {\tt y} and {\tt
z}, the task need only access the volatile page buffer.  The task updates the
page, leaving it dirty at the task's completion.  When the task completes, it
first copies the page from the page buffer into the non-volatile {\tt
pagesTemp} region before commit.  Then, \sys sets the commit bit, 
atomically commits the page back to its original memory location, and unsets
the commit bit.

@@

%One of the main objectives of \sys run-time is to restrict tasks to interface
%only with the volatile memory, since volatile memory access is relatively
%cheaper, as we already have shown experimentally in
%Section~\ref{section:background_task_computing} (Problem 1,
%Figure~\ref{fig:framEnergy}). This feature will also provide several advantages
%during task coalescing, as we will be dealing with in Section
%\ref{sec:task_coalescing}.

%\subsection{Intermittent Volatile-Only Memory Access}
%\label{sec:virtualization_problems}
%
%In order to enable volatile-only memory access, the persistent variables a task
%operates on should be brought from non-volatile memory to a fixed-sized
%\emph{volatile buffer} in the main memory. During its execution, the active
%task reads and modifies only the contents of this buffer. When the task is
%finished its execution, the modified variables, i.e. memory locations, in this
%buffer should be committed back to their original locations in non-volatile
%memory. We list three fundamental problems proving that this is however a
%non-trivial task. 
%
%
\subsection{Managing paging overhead}
The efficiency of \sys's paging scheme relies on the fact that data in pages
are contiguous, which makes them amenable to being manipulated by hardware
assisted DMA bulk copy operations. To justify this design choice we
experimentally evaluated the efficiency of using DMA block copies, compared to
explicit, software copy loops.  We copied blocks of data of different size
between memory regions on a WISP~\cite{wisp} using a hand-written loop and
using DMA.  Figure~\ref{fig:dmaTimeEnergy} shows that for moving large blocks
of data like a page, hardware DMA is considerably better than software in both
run time and energy efficiency.  \sys's design moves data in pages, not
individual words, leveraging the increased efficiency of DMA data movement.
%\paragraph{Problem 1---Runtime Access Overhead:} The persistent variables read and/or modified by a task might not be known in advance due to the dynamic program flow. Therefore, the volatile buffer should keep track of the persistent variables at runtime, as presented in Figure~\ref{fig:volatile-buffer}. When a persistent variable is read or written (refer to steps 1 and 3--5 in Figure~\ref{fig:volatile-buffer}) first the volatile copy of the corresponding variable is \emph{searched} in this buffer. If it is found, the read or write operation is performed on the volatile copy. Otherwise, the persistent variable is read from the non-volatile memory and inserted in this buffer for future operations (refer step 2 in Figure~\ref{fig:volatile-buffer}). Unfortunately, we observed that \emph{search--bring} operations at each access to the persistent variables introduce considerable overhead (refer to Figure~\ref{fig:dmaTimeEnergy} and pointer-based operation therein) and might nullify the benefit of task coalescing, a cores \sys feature.
%
\begin{figure}[t]
	\centering
	\subfloat[Time needed to transfer a block of data]{\includegraphics[width=0.49\columnwidth]{figures/dmaSize_time.eps} }
	\subfloat[Energy needed to transfer a block of data]{\includegraphics[width=0.49\columnwidth]{figures/energyConsumptionDMA_SW.eps}}
	\caption{Time and energy consumption of moving a block of data from SRAM to FRAM: CPU intervention versus Direct Memory Access (DMA).\todo{Explain HW/SW setup; make fonts larger, unify legends and axes}{Amjad}}
	\label{fig:dmaTimeEnergy}
\end{figure}
%
%\paragraph{Problem 2---Two Phase Commit Overhead:} For the consistency of the non-volatile memory, the volatile buffer should be committed at the end of each task by means of a \emph{two-phase commit}---refer Figure~\ref{fig:volatile-buffer} steps 6--8 for the first phase and 9--11 for the second step of the commit operation. However, two-phase commit is also expensive since the committed variables are not \emph{contiguous}---leading the fact that copying or moving of data between/within volatile and non-volatile memory should be always performed selectively with CPU intervention. Non-selective block memory copy operations using Direct Memory Access (DMA) that eliminates CPU intervention are considerably more efficient, refer again to Figure~\ref{fig:dmaTimeEnergy} for a comparison, but infeasible in this case.
%
%\paragraph{Problem 3---Memory Fit Constraints:} Persistent variables can be allocated \emph{contiguously} in non-volatile memory with compiler support. In order to eliminate run-time buffer \emph{search} overhead (by not relying on specialized hardware such as e.g. in~\cite{hicks_isca_2017}), \emph{all} persistent variables can be brought from non-volatile memory to the volatile buffer at system start-up---very efficiently and faster thanks to DMA for \emph{block} memory-memory copy/move operations. Two-phase commit is also very efficient in this case since DMA can be utilized to move volatile buffer as a block to the corresponding non-volatile locations. However, bringing all persistent variables from non-volatile memory to the volatile buffer suffers from the \emph{memory fit problem}: the volatile memory is a scarce resource and may not be large enough to hold all persistent variables.

%\todo{Benefits of the paging to the task coalescing should be mentioned in the next section}{Amjad/Przemek/Sinan}

>>>>

\subsection{Paged Non-Volatile Virtual Memory}

\sys uses a page-based virtual memory scheme that allows tasks to primarily use SRAM instead of frequently accessing FRAM. The paging scheme partitions the non-volatile memory address space into pages of {\em protected data}. As they are accessed by a task, pages of protected data are swapped into a {\em page buffer} that is stored in volatile memory and holds a single page. When a modified page is swapped out, it is buffered in a dedicated commit buffer for later commit. \sys uses two-phase commit to write buffered, modified pages back to their original memory locations atomically, ensuring data consistency even with power failures. \sys's page-based memory model allows moving large blocks of data between non-volatile and volatile memory, which \sys does efficiently using common,
hardware DMA block-copy operations that have higher bandwidth and avoid fixed, 
per-variable data movement costs. \sys uses efficient DMA block copies when paging data in and out, as well as during two-phase commit. \sys is unlike prior work~\cite{chain,alpaca} that moves one variable at a time, prohibiting efficient, DMA bulk transfers. Moreover, \sys's paging and buffering scheme is a key enabler of task coalescing, which we describe in Section~\ref{sec:task_coalescing}. Paging simplifies coalescing because deferring a commit of paged data at a task boundary entails leaving paged data buffered only, not requiring additional state management logic at task boundaries.

@@

\sys virtualizes the protected, non-volatile data that a task can access through a modified memory interface. The interface includes two operations: \texttt{RVAR(var)}, which reads the value of the memory location {\tt var} and \texttt{WVAR(var,val)}, which writes the value {\tt val} into the memory location {\tt var}. {\tt RVAR} and {\tt WVAR} operations translate a variable's physical address in non-volatile memory and translates it into a \emph{virtual address} in the page buffer \texttt{pageBuf} in volatile memory.  In \sys, a virtual address is composed of a \emph{page tag} and a \emph{page offset}. The page tag identifies in which page of non-volatile memory the memory location resides and
the page offset identifies a particular byte of that page. How an $a$-bit memory address decomposes into a tag and an offset depends on the number (and size) of each page of protected memory. To address {\tt p} pages of size {\tt s}, a tag comprises $log(p)$ bits, the offset comprises $log(s)$ bits and $log(p) + log(s) = a$. For example, in a 16-bit address space (i.e., 16\,kB of protected memory) with 512 pages of 128 bytes each, a
page's tag is the high-order nine bits of the address and a byte's page offset
is the remaining 7 bits.

@@

After address translation, \sys attempts to access the protected variable's location in the volatile paging buffer. \sys keeps track of the page tag for the page currently resident in the paging buffer in {\tt CrntPagTag}, which is stored in non-volatile memory. When a task accesses a variable, it compares the variable's page tag to {\tt CrntPagTag}.  If an accessed variable's page is equal to {\tt CrntPagTag} then the accessed variable's page is resident in the page buffer. If the accessed variable's page tag is not equal to {\tt CrntPagTag} then the variable's page is not resident in the page buffer. If the accessed variable's page is resident in the page buffer, an executing memory operation indexes the page buffer using the variable's page offset and manipulates that byte (i.e., either reading or writing). If the accessed variable's page is not resident in the page buffer, the operation incurs a {\em page fault}. At a page fault, \sys must swap out the
page that is resident int he page buffer and swap in the the variable's page.

Algorithm~\ref{algo:rwar} shows pseudo-code of \texttt{RVAR} and we omit a listing for {\tt WVAR} as it is very similar except it writes the memory location and returns nothing. If the page tag of the variable \texttt{var} and \texttt{CrntPagTag} are not equal (e.g. Lines 2--3), there is a page fault that requires a new page to be swapped in to the page buffer.

@@

A page fault on a memory access requires \sys to swap out whatever page of data is in the page buffer, preserving updates made to that page, and to copy the accessed page of protected data into the page buffer. Algorithm~\ref{algo:pagefault} shows the how \sys handles a page fault. If both the page being swapped out and the page being swapped in are not {\em dirty}---which means that they do not contain bytes modified by the executing
task---then handling a page fault simply requires copying the accessed page into the page buffer. If either the {\em resident} page being swapped out or the {\em accessed} page being swapped in is dirty, \sys must take extra steps when handling a page fault.

@@

\paragraph{Dirty accessed page.} If an accessed page is modified, \sys should swap it into the page buffer from its location in {\tt pagesTemp}, rather than swapping it in from its original location in memory. As Algorithm~\ref{algo:pagefault} shows in Line 3, \sys checks if the requested page was previously modified by the executing task. To track whether a page is modified, \sys stores a dirty bit with the page that is set when a task
modifies the page. If an accessed page's dirty bit is set, the page must be swapped in from \texttt{pagesTemp} rather than its original memory location (Line 4), furnishing the task with the page's modified values. If the accessed page is not dirty, \sys swaps it in from its original location in memory (Line 6). After swapping the page in, \sys sets \texttt{CrntPagTag} to the tag of the accessed page (Line 7). \sys uses DMA block copies for all page swapping operations, leveraging the efficiency of hardware support.

\subsubsection{Committing a Page}

@@

\subsection{Paging Example}

\todo{Redo this figure to actually reflect paging -- right now it isn't clear if x,y,z are in a page or not, it isn't clear when the page gets copied to pagesTemp, it isn't clear what state we add to the system to help with commit, etc (i.e., dirty bits, commit bit).  Make this figure's labels consistent with the text: pagesTemp, RVAR, WVAR, explicit
commit in task, Page Buffer}{whoever}

Figure~\ref{fig:volatile-buffer} illustrates the execution of a simple task using \sys's paging mechanism. The task manipulates protected variables {\tt x}, {\tt y}, and {\tt z}, which are in the same page. On accessing {\tt x}, its page is swapped into the page buffer. On accessing {\tt y} and {\tt z}, the task need only access the volatile page buffer. The task updates the page, leaving it dirty at the task's completion. When the task completes, it first copies the page from the page buffer into the non-volatile {\tt
pagesTemp} region before commit. Then, \sys sets the commit bit, atomically commits the page back to its original memory location, and unsets the commit bit.

@@

\subsection{Managing Paging Overhead}

The efficiency of \sys's paging scheme relies on the fact that data in pages are contiguous, which makes them amenable to being manipulated by hardware assisted DMA bulk copy operations. To justify this design choice we experimentally evaluated the efficiency of using DMA block copies, compared to explicit, software copy loops. We copied blocks of data of different size between memory regions on a WISP~\cite{wisp} using a hand-written loop and using DMA (refer to Section~\ref{sec:results_hardware} for hardware setup details). Figure~\ref{fig:dmaTimeEnergy} shows that for moving large blocks of data like a page, hardware DMA is considerably better than software in both run time and energy efficiency. \sys's design moves data in pages, not individual words, leveraging the increased efficiency of DMA data movement.

\begin{figure}[t]
	\centering
	\subfloat[Time needed to transfer a block of data]{\includegraphics[width=0.49\columnwidth]{figures/dmaSize_time.eps} }
	\subfloat[Energy needed to transfer a block of data]{\includegraphics[width=0.49\columnwidth]{figures/energyConsumptionDMA_SW.eps}}
	\caption{Time and energy consumption of moving a block of data from SRAM to FRAM: CPU intervention versus Direct Memory Access (DMA).\todo{Explain HW/SW setup; make fonts larger, unify legends and axes}{Amjad}}
	\label{fig:dmaTimeEnergy}
\end{figure}
